---
title: "Fertility Rate Estimates"
output:
  html_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

source("R/functions/plots.R")

```

```{r}
# Read in raw and smooth fertility rates
asfr_lad_smooth <- readRDS("data/processed/smooth_asfr_lad_agg_cy.rds")
asfr_lad_raw <- readRDS("data/processed/raw_asfr_lad_agg_cy.rds")

# Create total fertility rates
total_fertility_rates <- readRDS("data/processed/tfr_lad_agg_cy.rds")

```
<br />

### Introduction to fertility estimates \

Fertility refers to the number of live births within an individual or group, influenced by a combination of biological, social, cultural, and economic factors. At the GLA, we produce age-specific and total fertility rate estimates for each local authority and higher level geography in England and Wales. These estimates are a key input into our annual population estimates. \

More information on how age-specific and total fertility rates are calculate and project methodology can be found in our technical notes below.

The codebase for this project is [available on GitHub](https://github.com/Greater-London-Authority/fertility_rate_estimation/tree/main). All the requirements and information necessary to run the analysis can be found in the README file.

### The data \

The data used to calculate fertility rate estimates are:

- ONS calendar year births by age of mother from 1993
- ONS mid-year population estimates from 1991 - 2000
- GLA estimates components of change modelled backseries from 2000

After processing the data, we calculate age-specific fertility rates for each local authority and higher level geographies in England and Wales. Finally, we apply a smoothing function to the fertility rates to reduce noise.

Age-specific fertility rates are then added to obtain total fertility rates. \

The snapshot below is extracted from the age-specific fertility rates output:

```{r, echo=FALSE, message=FALSE, results="asis", warning=FALSE}

top_row <- asfr_lad_smooth %>% filter(gss_code == "E09000028" & year == 1995) %>% head(4)
bottom_row <- asfr_lad_smooth %>% filter(gss_code == "E09000028" & year == 2023) %>% tail(2)
top_row[] <- lapply(top_row, as.character)
bottom_row[] <- lapply(bottom_row, as.character)

bottom_row[1,] <- "..."
top_row[1,] <- "..."
table <- bind_rows(top_row, bottom_row)

table %>% kableExtra::kable(caption = "smooth_asfr_lad_agg_cy.rds") %>% 
  kableExtra::kable_styling(font_size = 13)

```

From the age-specific fertility rates, we calculate total fertility rates:

```{r, echo=FALSE, message=FALSE, results="asis", warning=FALSE}

top_row <- total_fertility_rates %>% filter(gss_code == "E09000028" & year %in% c(1993, 1994, 1995, 1996))
bottom_row <- total_fertility_rates %>% filter(gss_code == "E09000028" & year %in% c(2022, 2023))
top_row[] <- lapply(top_row, as.character)
bottom_row[] <- lapply(bottom_row, as.character)

bottom_row[1,] <- "..."
top_row[1,] <- "..."
table <- bind_rows(top_row, bottom_row)

table %>% kableExtra::kable(caption = "tfr_lad_agg_cy.csv") %>% 
  kableExtra::kable_styling(font_size = 13)


```

### Analysis \

The output can then be plotted in various ways. If users are familiar with R, our [GitHub repository]((https://github.com/Greater-London-Authority/fertility_rate_estimation/tree/main)) has all the functionality necessary for users to plot age-specific and total fertility rates across local authorities and periods of interest.


```{r dpi=300}
lad_codes <- c("E09000012", "E09000027", "E09000009", "E09000011")
years <- c(2003, 2013, 2023)

plot_asfr_multiple_areas_and_years(asfr_data = asfr_lad_smooth, 
                                   lad_codes = lad_codes, 
                                   years = years, 
                                   ncols = 3, 
                                   label_group = "gss_name", 
                                   title = "Smooth Age-Specific Fertility Rates")

```


```{r dpi=300}

start_year <- min(total_fertility_rates$year)
end_year <- max(total_fertility_rates$year)

plot_tfr_selected_areas(tfr_data = total_fertility_rates, 
                        lad_code = lad_codes, 
                        years = c(start_year:end_year))

```


