---
title: "Tempo Adjusted Fertility Rate Estimates"
author: "Demography - Greater London Authority"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: html_document
bibliography: references.bib
link-citations: true
---

### Introduction to tempo adjusted fertility estimates \

The calculation of Total Fertility Rates (TFR) reflects the interaction of 
two key components: **tempo** (the timing of childbearing) and **quantum** 
(the overall level of childbearing). 

When the mean age at childbearing shifts, standards measure of fertility can
misrepresent the underlying fertility behaviour. For example, if a
cohort of women postpones childbearing, the completed number of children 
they eventually have may remain unchanged. However, this delay in timing 
(a tempo effect) can temporarily depress the observed TFR, creating the 
impression of a decline in fertility levels (a quantum effect). In this way,
changes in timing may be misinterpreted as changes in fertility levels.

In England, the age at which women have children has increased steadily since 
the early 1990s - a common trend across may developed countries. This pattern 
is also evident at the Local Authority level, as illustrated the graph below:

```{r results='hide'}
knitr::opts_chunk$set(echo = FALSE)

source("R/functions/tempo_adjusted_fertility_rates.R")
source("R/functions/plots.R")

file_path <- "data/processed/adjusted_tfr_my.rds"
asfr_lad_smooth <- readRDS("data/processed/smooth_asfr_lad_agg_cy.rds")

if (!file.exists(file_path)) {
  adjusted_tfr <- calculate_tempo_adj_tfr(
  "data/processed/smooth_asfr_lad_agg_cy.rds")
  saveRDS(adjusted_tfr, file_path)
} else {
  adjusted_tfr <- readRDS(file_path)
}
```



```{r results='hide'}
lad_codes <- c("E09000007", "E09000020", "E09000026", "E12000007", "E92000001")

adjusted_tfr |>
  filter(gss_code %in% lad_codes) |>
  ggplot(aes(x = year, y = mac, colour = gss_name)) +
  theme_gla(base_size = 10) +
  geom_line(linewidth = 1, alpha = 0.7) +
  scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(25, NA)) +
  labs(
    title = "Mean childbearing age",
    subtitle = "",
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )

```

### How tempo-adjusted fertility rates are calculated \

The first tempo adjustment was proposed by Bongaarts and Feeney (1998):

$$\text{adjTFR}(t)_{i} = \frac {TFR_{i}(t)}{1-r_{i}(t)}$$
Where: \
$r_{i}$ the change in the mean age at childbearing (MAC) of birth order {i} \

Because data on birth order by Local Authority in England and Wales is not 
currently available, we calculate the adjusted rates by using the TFR and mean
age at childbearing for each Local Authority and year.

Calculating tempo-adjusted TFR without parity-specific data relies on the
assumption that fertility is below replacement level, that changes in the mean 
age at childbearing occur uniformly across birth orders, and that the annual 
shift in mean age does not exceed about 0.2 years. However, without birth-order 
data, it is impossible to verify whether these assumptions hold in practice. 

As a result, the adjusted TFR may be biased if different parities experience 
different timing shifts or if structural changes in fertility are occurring 
simultaneously.

This is particularly relevant in London, where high levels of migration can 
alter the parity composition, age structure, and fertility behaviour of 
the population, making it unlikely that the uniform timing-shift assumption 
holds.



```{r}

plot_asfr_multiple_areas_and_years(asfr_data = asfr_lad_smooth, 
                                   lad_codes = c(
                                     "E09000018", "E09000005", "E09000025", "E09000012"),
                                   years = c(2001, 2002, 2003), 
                                   ncols = 6, 
                                   label_group = "year")


```




```{r results='hide'}

adjusted_tfr |>
  filter(gss_name == "England") |>
  ggplot(aes(x = year)) +
  theme_gla(base_size = 10) +
  geom_line(aes(y = tfr, linetype = "tfr"), linewidth = 1, alpha = 0.7) +
  geom_line(
    aes(y = adj_tfr, linetype = "adj_tfr"),
    linewidth = 1,
    alpha = 0.7
  ) +
  scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_linetype_manual(
    values = c("tfr" = "solid", "adj_tfr" = "dotted"),
    labels = c("tfr" = "Observed TFR", "adj_tfr" = "Adjusted TFR"),
    name = NULL
  ) +
  labs(
    title = "Adjusted TFR",
    subtitle = location,
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )

```


