---
title: "Tempo Adjusted Fertility Rate Estimates"
author: "Demography - Greater London Authority"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: html_document
bibliography: adjusted_tfr.bib
link-citations: true
---

```{css, echo=FALSE}
div {
    text-align: justify;
    margin-top: 10px;
    width: 600px;
}
```

```{r, include=FALSE}

source("R/functions/tempo_adjusted_fertility_rates.R")
source("R/functions/plots.R")

file_path <- "data/processed/adjusted_tfr_my.rds"
asfr_lad_smooth <- readRDS("data/processed/smooth_asfr_lad_agg_cy.rds")

if (!file.exists(file_path)) {
  adjusted_tfr <- calculate_tempo_adj_tfr(
  "data/processed/smooth_asfr_lad_agg_cy.rds")
  saveRDS(adjusted_tfr, file_path)
} else {
  adjusted_tfr <- readRDS(file_path)
}
```

### Introduction to tempo adjusted fertility estimates \

The calculation of Total Fertility Rates (TFR) reflects the interaction of 
two key components: **tempo** (the timing of childbearing) and **quantum** 
(the overall level of childbearing).

When the mean age at childbearing shifts, standards measure of fertility can
misrepresent the underlying fertility behaviour. For example, if a
cohort of women postpones childbearing, the completed number of children 
they eventually have may remain unchanged. However, this delay in timing 
(a tempo effect) can temporarily depress the observed TFR, creating the 
impression of a decline in fertility levels (a quantum effect). In this way,
changes in timing may be misinterpreted as changes in fertility levels.

In England, the age at which women have children has increased steadily since 
the mid-1970s - a common trend across may developed countries. This pattern 
is also evident at the Local Authority level, as illustrated the graph below:


```{r, echo=FALSE}
lad_codes <- c("E09000007", "E09000020", "E09000026", "E12000007", "E92000001")
years <- c(min(adjusted_tfr$year):max(adjusted_tfr$year))

adjusted_tfr |>
  filter(gss_code %in% lad_codes) |>
  ggplot(aes(x = year, y = mac, colour = gss_name)) +
  theme_gla(base_size = 10) +
  geom_line(linewidth = 1, alpha = 0.7) +
  scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(25, NA)) +
  labs(
    title = "Mean childbearing age",
    subtitle = "",
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )

```

### How tempo-adjusted fertility rates are calculated \

The first tempo adjustment was proposed by Bongaarts and Feeney [@bongaarts_quantum_1998]:

$$\text{adjTFR}(t)_{i} = \frac {TFR_{i}(t)}{1-r_{i}(t)}$$
Where: \
$r_{i}$ the change in the mean age at childbearing (MAC) of birth order ${i}$ \

Because data on birth order by Local Authority in England and Wales is not 
currently available, we calculate the adjusted rates by using the TFR and mean
age at childbearing for each Local Authority at a given year.

According to Vienne Institute of Demography [@potancokova_estimating_2008], 
calculating tempo-adjusted TFR without  parity-specific data relies on three 
assumptions: 1) fertility is below replacement level 2) changes in the mean age 
at childbearing occur uniformly across birth orders, 
and 3) the annual shift in mean age does not exceed about 0.2 years. 

As a result, the adjusted TFR may be biased if different parities experience 
different timing shifts or if structural changes in fertility are occurring 
simultaneously. 

This is particularly relevant in London, where high levels of migration can 
alter the parity composition, age structure, and fertility behaviour of 
the population, making it unlikely that the assumptions two and three hold.

### Preliminary discussion

The GLA Demography team calculated adjusted rates for all regions, (ITL level 2) subregions, and Local Authority Districts 
in England and Wales. These results are presented as preliminary research outputs only.
team still trying to make sense of the results and how they can
help us understand fertility rates in London. 

Adjusted fertility rates aimed to remove the temporary tempo distortions 
caused by an increase in the mean age at childbearning. As expected, the 
adjusted TFR is above the standard TFR for England and London, 
as demonstrated below:

```{r}

location <- c("England", "London")

adjusted_tfr |>
  dplyr::filter(gss_name == "England") |>
  ggplot(aes(x = year)) +
  theme_gla(base_size = 10) +
  geom_line(aes(y = tfr, linetype = "tfr"), linewidth = 1, alpha = 0.7) +
  geom_line(
    aes(y = adj_tfr, linetype = "adj_tfr"),
    linewidth = 1,
    alpha = 0.7
  ) +
  geom_line(
    data = adjusted_tfr |>
    dplyr::filter(gss_name == "London") |> 
    dplyr::rename(adj_tfr_london = adj_tfr), mapping = aes(y = adj_tfr_london, linetype = "adj_tfr_london"),
    linewidth = 1,
    alpha = 0.7
  ) +
   geom_line(
    data = adjusted_tfr |>
    dplyr::filter(gss_name == "London") |> 
    dplyr::rename(tfr_london = tfr), mapping = aes(y = tfr_london, linetype = "tfr_london"),
    linewidth = 1,
    alpha = 0.7
  ) +
  scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_linetype_manual(
    values = c("tfr" = "solid", "adj_tfr" = "dotted", "adj_tfr_london" = "dotdash", "tfr_london" = "longdash"),
    labels = c("tfr" = "TFR England", "adj_tfr" = "Adjusted TFR England", "adj_tfr_london" = "Adjusted TFR London", "tfr_london" = "TFR London"),
    name = NULL
  ) +
  labs(
    title = "TFR and Adjusted TFR",
    subtitle = "England and London",
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )

```

Below, we've outlined the few points of consideration: \

**1. Hypersensitive to changes in mean age at childbearing (MAC)** \
Even though the MAC graph above shows that MAC has been increasing linearly, 
the rate of change varies significantly as demonstrated in the graph below: \

```{r}

ggplot(data = adjusted_tfr |>
           dplyr::filter(gss_name == "Camden"), 
       mapping = aes(y=delta_mac, x= years, linetype = "delta_mac")) +
    geom_line(linewidth = 1,
    alpha = 0.7) +
    geom_line(data = adjusted_tfr |>
           dplyr::filter(gss_name == "London") |> 
            dplyr::rename(delta_mac_london = delta_mac), 
       mapping = aes(y=delta_mac_london, linetype = "delta_mac_london"), 
        linewidth = 1,
    alpha = 0.7) +
    theme_gla(base_size = 10) +
    scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(-0.7, 1.0)) +
   scale_linetype_manual(
    values = c("delta_mac_london" = "solid", "delta_mac" = "dotted"),
    labels = c("delta_mac" = "\U0394MAC Camden", "delta_mac_london" = "\U0394MAC London"),
    name = NULL
  ) +
  labs(
    title = "Change in mean age at childbearing (MAC)",
    subtitle = "Camden and London",
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )
  
```

Both graphs shows how MAC and adjusted TFR follow a similar pattern. \

When MAC increases rapidly the adjusted fertility rates becomes inflated. 
For example, MAC rose substantially from 2005, falling in 2008, before rising 
sharply again in 2009/10. Adjusted TFR peaks in London in 2012, when we observe 
the largest increase in MAC - 0.50, double of the average change for the 
period (0.24). \ 

This is particularly problematic at the Local Authority level, where $\Delta{MAC}$
is highly volatile. In Camden for example, from 1994 to 2022 the YoY change in 
MAC varied from a high of 0.97 and a low of -0.28. \

```{r}

adjusted_tfr |>
  dplyr::filter(gss_name == "Camden") |>
  ggplot(aes(x = year)) +
  theme_gla(base_size = 10) +
  geom_line(aes(y = tfr, linetype = "tfr"), linewidth = 1, alpha = 0.7) +
  geom_line(
    aes(y = adj_tfr, linetype = "adj_tfr"),
    linewidth = 1,
    alpha = 0.7
  ) +
  scale_x_continuous(breaks = years, guide = guide_axis(angle = 60)) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_linetype_manual(
    values = c("tfr" = "solid", "adj_tfr" = "dotted"),
    labels = c("tfr" = "TFR", "adj_tfr" = "Adjusted TFR"),
    name = NULL
  ) +
  labs(
    title = "Camden TFR and Adjusted TFR",
    caption = "Source: GLA modelled rates based on ONS births by age of mother"
  )
```


**2. Lack of birth data by parity at the Local Authority level** \
The ONS does not publish birth data by parity at the Local Authority level, which
limits our ability to validate assumptions and verify trends. For example, we
could check whether MAC by birth order is less volatile and if higher-order 
births are more/less common in some boroughs.


**3. Hard to separate tempo from reduction in family size** \
Changes in fertility rates are assumed to be solely tempo changes. We cannot
be sure that all changes can be attributed to tempo dynamics, it is plausible
that postponment is accompanied by a reduction in family size. 

Data from the Netherlands 1950-97 shows how the assumption that women of all ages 
bearing children defer or advance their births in the same manner does is not 
supported by evidence [@van_imhoff_quantum_2000]. Period tempo changes depend
on cohort. Similarly, this assumption is also violated at small areas.


```{r, echo=FALSE}

plot_asfr_multiple_areas_and_years(asfr_data = asfr_lad_smooth, 
                                   lad_codes = c(
                                     "E09000018", "E09000005", "E09000025", "E09000012"),
                                   years = c(2001, 2002, 2003), 
                                   ncols = 6, 
                                   label_group = "year")


```


### Final remarks \

### Bibliography \
